Evaluation of Cache-Oblivious Data Structures [draft-1]

### 2008-05-26 paper/paper.tex
Fixed some typos. Moved some bibiography entries around.

\begin{table}
\begin{center}
\begin{tabular}{ l l l l }
\hline
\textbf{Type} & \textbf{Total Size} & \textbf{Block Size} & \textbf{Access Time} \\
\hline
Registers  &   32 bytes & 4 bytes & 1 cycle \\
L1 Cache   &   32 KB    & 64 bytes & 3 cyles \\
L2 Cache   &    1 MB    & 64 bytes & 9 cycles \\
RAM        &   ~2 GB    & 1 KB     & 50-100 cycles \\
Disk       & ~300 GB    & 4 KB     & 5,000-10,000 cycles \\
\hline
\end{tabular}
\caption{Sizes and access times of various types of storage}
\label{tab-memhier}
\end{center}
\end{table}

To conclude: the memory model used in real computer systems is quite a bit more complex than the uniform memory model assumed. Given this reality, the performance of many existing algorithms and data structures can be improved by taking the existence of a memory hierarchy into account. This has prompted research into new memory models that are more realistic. We will consider three classes of data structures and algorithms, depending on the assumptions that are made about the environment in their definition:
\begin{list}{}{}
\item \textbf{Cache-unaware} data structures and algorithms are designed with the traditional uniform memory model in mind, without considering the existance of a memory hierarchy, and may therefore not perform optimally in practice.
\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and algorithms are designed to perform optimally in the external memory model (described below) but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model (explained below) which does not allow para\-metri\-zation with properties of the cache layout.
\end{list}

Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice. Altough some research has been done on cache-oblivious data structures, they are seldom used in practice. The main contribution of this paper is to give some insight in the practical merit of cache-oblivious data structures.

@@

\subsection{Cache-Oblivious Memory Model}
A different approach to generalizing the external memory model was taken by Prokop \cite{prokop1999coa} who proposed the cache-oblivious model, which consists of an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory. Data is transferred between the two in aligned data blocks of size B. In contrast with the hierarchical memory model, the application does not have explicit control over the transfering of blocks between the two memories. Instead, it is assumed that an optimal cache manager exists which minimizes the number of block transfers over the execution the program. Additionally (and in contrast with the external memory model) the values of parameters like M and B are not known to the application, so they cannot be used explicitly when defining algorithms and datastructures. Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).

@@

Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize seperate-chaining hash tables for cache-efficiency by storing the linked lists that contain the contents for a bucket in contiguous memory. Their experiments show a performance gain over traditional methods, especially when the hash table is heavily loaded.

@@

From the pseudo-code it is clear that the state space search algorithm does not
accomplish much by itself; instead, the real work is done by the successor
function and the queue and the set datastructures. Queues can easily be
implemented efficiently (adding and removing states takes O(1) time).
The efficiency of the successor function depends on the execution model used,
but is typically linear in the number of states produced. In practice,
therefore, the bottleneck of the algorithm is the set data structure.

@@

The use of memory mapped files means that the operating system is in charge of
when and which pages are transferred between main memory and disk storage; the
application assumes no control over this policy. This is in accordance with the
cache-oblivious model, that assumes no control over the cache is possible.
Additionally, memory mapped files bypass the C library's allocation functions
and give the implementor complete control over how data is ordered in (virtual)
memory.

\subsubsection{Hash Table Implementation}
In its simplest form, a hash table consists of an
index array (the \emph{index}) with a fixed number of slots. A \emph{hash
function} is used to map key values onto slot numbers. If the slot for a value
is empty, it may be inserted there. Queries for the existance of an element in
the hash table similarly see if the queried value is stored at its designated
slot.

@@

B-tree pages are ordered in a tree structure. Values in a page are stored
in order, and pointers are added that originate between a pair of adjacent
values, and point to a subtree containing only values that fall in between those
values. The resulting tree is typically very shallow, which is benefical,
as the number of pages that need to be retrieved (worst case) is equal to
the height of the tree.

@@

B-trees can easily support the insertion of variable-length values, as long
as individual values fit in the size of a single page. However, there is no
obvious way to store values that are larger than the size of a single page.
Production-quality implementations handle this case separately. Our our
implementation does not support this requires stored values to
be smaller than the pagesize. Additionally, deletion of items has not been
implemented.

@@

In the tree, the leaf nodes (when numbered from left to right) store the same values
as the sparse array (including empty values).
Interior nodes store the maximum value of their two children (or an empty value,
if both children store an empty value).
This tree can be used in a similar way as a binary search tree: by examining the
two children of a node, we can determine wether a searched-for value lies in the
left or right subtree.

@@

The following models are used for benchmarking:
\begin{itemize}
\item\textbf{Eratosthenes} is a parallel implementation of the sieve
of Eratosthenes which is used to find prime numbers upto a given maximum.
It has a single parameter $N$: only integers from 2 to $N$ (exclusive) are
tested. A new process is created for every prime number found, and as a result
the state space can only by exhaustively searched for relatively small values
of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
increases during execution.

@@

TODO: describe parameters of data structures
Explain results for B-tree (higher page size packs more data), hash table (larger index, less collissions, cache miss anyway)

@@

\section{Conclusion}
The experimental results clearly show that in terms of both memory usage and
execution time, the cache-oblivious data structure proposed by Bender, Duan,
Iacono and Wu is outperformed by traditional data structures in our test scenario.
The advantage of more cache-friendly behaviour does not appear to be large enough
to compensate for the increased complexity of the data structure, which results in
high ermemory requirements and increased computational overhead.
If the data structure does have asymptotic performance benefits, then realistic
work loads on current hardware systems are not enough to reveal this.
These findings are consistent with earlier results, such as obtained by
Rahman, Cole and Raman, and Olsen and Skov (see Related Work above).

>>>>

\begin{table}
\begin{center}
\begin{tabular}{ l l l l }
\hline
\textbf{Type} & \textbf{Total Size} & \textbf{Block Size} & \textbf{Access Time} \\
\hline
Registers  &   32 bytes & 4 bytes & 1 cycle \\
L1 Cache   &   32 KB    & 64 bytes & 3 cycles \\
L2 Cache   &    1 MB    & 64 bytes & 9 cycles \\
RAM        &   ~2 GB    & 1 KB     & 50-100 cycles \\
Disk       & ~300 GB    & 4 KB     & 5,000-10,000 cycles \\
\hline
\end{tabular}
\caption{Sizes and access times of various types of storage}
\label{tab-memhier}
\end{center}
\end{table}

To conclude: the memory model used in real computer systems is quite a bit more complex than the uniform memory model assumed. Given this reality, the performance of many existing algorithms and data structures can be improved by taking the existence of a memory hierarchy into account. This has prompted research into new memory models that are more realistic. We will consider three classes of data structures and algorithms, depending on the assumptions that are made about the environment in their definition:
\begin{list}{}{}
\item \textbf{Cache-unaware} data structures and algorithms are designed with the traditional uniform memory model in mind, without considering the existence of a memory hierarchy, and may therefore not perform optimally in practice.
\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and algorithms are designed to perform optimally in the external memory model (described below) but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model (explained below) which does not allow para\-metri\-zation with properties of the cache layout.
\end{list}

Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice. Although some research has been done on cache-oblivious data structures, they are seldom used in practice. The main contribution of this paper is to give some insight in the practical merit of cache-oblivious data structures.

@@

\subsection{Cache-Oblivious Memory Model}
A different approach to generalizing the external memory model was taken by Prokop \cite{prokop1999coa} who proposed the cache-oblivious model, which consists of an infinitely large external memory and an internal memory of size M which operates as a cache for the external memory. Data is transferred between the two in aligned data blocks of size B. In contrast with the hierarchical memory model, the application does not have explicit control over the transferring of blocks between the two memories. Instead, it is assumed that an optimal cache manager exists which minimizes the number of block transfers over the execution the program. Additionally (and in contrast with the external memory model) the values of parameters like M and B are not known to the application, so they cannot be used explicitly when defining algorithms and data structures. Of course, analysis in terms of memory transfers does involve these parameters, so the number of memory transfers performed is still a function of M and B (and other parameters relevant to the problem).

@@

Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize separate-chaining hash tables for cache-efficiency by storing the linked lists that contain the contents for a bucket in contiguous memory. Their experiments show a performance gain over traditional methods, especially when the hash table is heavily loaded.

@@

From the pseudo-code it is clear that the state space search algorithm does not
accomplish much by itself; instead, the real work is done by the successor
function and the queue and the set data structures. Queues can easily be
implemented efficiently (adding and removing states takes O(1) time).
The efficiency of the successor function depends on the execution model used,
but is typically linear in the number of states produced. In practice,
therefore, the bottleneck of the algorithm is the set data structure.

@@

The use of memory mapped files means that the operating system is in charge of
when and which pages are transferred between main memory and disk storage; the
application assumes no control over this policy. This is in accordance with the
cache-oblivious model, that assumes no control over the cache is possible.
Additionally, memory mapped files bypass the C library's allocation functions
and give the implementer complete control over how data is ordered in (virtual)
memory.

\subsubsection{Hash Table Implementation}
In its simplest form, a hash table consists of an
index array (the \emph{index}) with a fixed number of slots. A \emph{hash
function} is used to map key values onto slot numbers. If the slot for a value
is empty, it may be inserted there. Queries for the existence of an element in
the hash table similarly see if the queried value is stored at its designated
slot.

@@

B-tree pages are ordered in a tree structure. Values in a page are stored
in order, and pointers are added that originate between a pair of adjacent
values, and point to a subtree containing only values that fall in between those
values. The resulting tree is typically very shallow, which is beneficial,
as the number of pages that need to be retrieved (worst case) is equal to
the height of the tree.

@@

B-trees can easily support the insertion of variable-length values, as long
as individual values fit in the size of a single page. However, there is no
obvious way to store values that are larger than the size of a single page.
Production-quality implementations handle this case separately. Our our
implementation does not support this requires stored values to
be smaller than the page size. Additionally, deletion of items has not been
implemented.

@@

In the tree, the leaf nodes (when numbered from left to right) store the same values
as the sparse array (including empty values).
Interior nodes store the maximum value of their two children (or an empty value,
if both children store an empty value).
This tree can be used in a similar way as a binary search tree: by examining the
two children of a node, we can determine whether a searched-for value lies in the
left or right subtree.

@@

The following models are used for benchmarking:
\begin{itemize}
\item\textbf{Eratosthenes} is a parallel implementation of the sieve
of Eratosthenes which is used to find prime numbers up to a given maximum.
It has a single parameter $N$: only integers from 2 to $N$ (exclusive) are
tested. A new process is created for every prime number found, and as a result
the state space can only by exhaustively searched for relatively small values
of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
increases during execution.

@@

TODO: describe parameters of data structures
Explain results for B-tree (higher page size packs more data), hash table (larger index, less collisions, cache miss anyway)

@@

\section{Conclusion}
The experimental results clearly show that in terms of both memory usage and
execution time, the cache-oblivious data structure proposed by Bender, Duan,
Iacono and Wu is outperformed by traditional data structures in our test scenario.
The advantage of more cache-friendly behavior does not appear to be large enough
to compensate for the increased complexity of the data structure, which results in
high ermemory requirements and increased computational overhead.
If the data structure does have asymptotic performance benefits, then realistic
work loads on current hardware systems are not enough to reveal this.
These findings are consistent with earlier results, such as obtained by
Rahman, Cole and Raman, and Olsen and Skov (see Related Work above).
